---
title: Chp 5 - Exploring numerical data
subtitle: Boxplots, median, and IQR
format: revealjs
auto-stretch: false
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(openintro)
```

## In groups

-   Practice from last time: $\S$ 5.10, #8, 9, 11, 22

## `nycflights`

```{r}

glimpse(nycflights)
```

## Arrival delays at PDX

How would you describe the distribution?

```{r}
#| echo: false

pdx_flights <- nycflights |> filter(dest == "PDX")
lax_flights <- nycflights |> filter(dest == "LAX")
```

```{r}
ggplot(data = pdx_flights, aes(x = arr_delay)) +
  geom_histogram(binwidth = 10, fill = "plum4", color="white")
```

## Box plots

Another way to visualize the distribution for `arr_delay`.

```{r}
#| echo: true

ggplot(data = pdx_flights, aes(x = arr_delay)) +
  geom_boxplot()
```

## Box plots: median and quartiles

-   The *box* contains the middle 50% of the data points. The **median** splits the data in half.
-   The *whiskers* depict each of the remaining **quartiles**.

```{r, out.width = "80%"}
#| echo: false

ggplot(data = pdx_flights, aes(x = arr_delay)) +
    geom_boxplot()

```

## Box plots: interquartile range

-   The length of the box is the **interquartile range**, or **IQR**
-   $IQR = Q_3 - Q_1$

```{r, out.width = "85%"}
#| echo: false

ggplot(data = pdx_flights, aes(x = arr_delay)) +
    geom_boxplot()

```

## Box plots: interquartile range

```{r}
pdx_flights |>
  summarise(median_ad = median(arr_delay), 
            IQR_ad = IQR(arr_delay))
```

```{r, out.width = "80%"}
#| echo: false

ggplot(data = pdx_flights, aes(x = arr_delay)) +
    geom_boxplot()

```

## Outliers

We can remove outliers to inspect the bulk of the data more closely.

```{r, out.width = "75%"}
#| echo: true
ggplot(data = pdx_flights, aes(x = arr_delay)) +
  geom_boxplot(outliers = FALSE)
```

## Another Example

```{r}
ggplot(data = nycflights, aes(x = arr_delay)) +
  geom_histogram(binwidth = 10, fill = "plum4", color="white")
```

## Same Distribution
```{r}
ggplot(data = nycflights, aes(x = arr_delay)) +
  geom_boxplot()
```

## Same Distribution
```{r}
ggplot(data = nycflights, aes(x = arr_delay)) +
  geom_boxplot(outliers = FALSE)
```


## Outliers

-   A data point is generally considered an **outlier** when it is more than $1.5 \times IQR$ away from the 1st or 3rd quartile.

-   An outlier is not necessarily a mistake or a problem, just extreme relative to the rest of the data

## Practice

(1) Sketch a possible box plot for a data set that is strongly left-skewed with two low outliers.
(2) In what kind of distribution is the median smaller than the mean?
