---
title: Hypothesis Testing
subtitle: Chapter 11
format: revealjs
auto-stretch: false
author: ""
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(openintro)
library(tidymodels)
library(palmerpenguins)
library(ggridges)
```

## In groups

-   Discuss Homework: Chapter 11, #1, 2

## Remember the Penguins!

```{r}
#| echo: true
#| warning: false

ggplot(penguins, mapping = aes(x = body_mass_g)) +
  geom_histogram(binwidth=200, col="white", fill = "seagreen")
```

## Density plot (new)

```{r}
#| echo: true
#| warning: false

ggplot(penguins, mapping = aes(x = body_mass_g)) +
  geom_density(fill = "seagreen")
```

## Box plot

```{r}
#| echo: true
#| warning: false

ggplot(penguins, mapping = aes(x = body_mass_g)) +
  geom_boxplot(fill = "seagreen")
```

## Body mass *faceted* by `species`

```{r}
#| echo: true
#| warning: false
#| fig-width: 10
#| fig-asp: 0.5
#| fig-align: center

ggplot(penguins, mapping = aes(x = body_mass_g)) +
  geom_histogram(binwidth=200, col="white", fill = "seagreen") +
  facet_wrap(~species)
```

## Mulitple boxplots

```{r}
#| echo: true
#| warning: false

ggplot(penguins, mapping = aes(x = body_mass_g, fill = species)) +
  geom_boxplot()
```

## Overlay density plots (new!)

```{r}
#| warning: false
#| echo: true

ggplot(penguins, mapping = aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = 0.3)
```

## Ridge plot (new!)

```{r}
#| echo: true
#| warning: false

ggplot(penguins, aes(x = body_mass_g, y = species, fill = species)) +
  geom_density_ridges(alpha = 0.3) 
```

## Summary Statistics

```{r}
#| echo: true

penguins |>
  group_by(species) |>
  summarize( mean = mean(body_mass_g, na.rm = TRUE),
             sd = sd(body_mass_g, na.rm = TRUE)
  )
    
```

\

**Question:** Are the mean body weights of Adelie and Chinstraps *really* different? Or is the difference just due to random *variation*?

## Sample vs Population

-   Chapter 5: sample mean is (hopefully) an approxmiation of the population mean

-   Clearly the **sample means** are different (3701 vs. 3733)

-   But are the **population means** different?

::: fragment
### What do you think?
:::

## Hypothesis Testing

-   $H_0$ -- Null Hypothesis (no difference)

Mean body weight of Adelie and Chinstrap penguins are the same.

-   $H_1$ -- Alternative Hypothesis (there is a difference)

Mean body weight of Adelie and Chinstrap penguins are different.

## Coin Flipping

If you flip a fair coin 100 times, how many times will it come up **heads**?

[Do a simulation](https://www.rossmanchance.com/applets/2021/oneprop/OneProp.htm)

## Gender Discrimination Study

Textbook [Section 11.1](https://openintro-ims.netlify.app/foundations-randomization#sec-caseStudySexDiscrimination)
