---
title: Confidence Intervals with Bootstrapping
subtitle: Chapter 12
format: revealjs
auto-stretch: false
author: ""
echo: false
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(openintro)
library(infer)

```

```{r}
#| echo: false
#| message: false

organ_donor <- read_csv("data/organ-donor.csv")
```

## In groups

-   Discuss Homework: Chapter 12, #1, 3, 7

## Confidence Intervals

\

### Goal is to estimate a population parameter

$p$ -- true value of a parameter (usually unknown)

$\hat {p}$ -- sample statistic

\

How **confident** can we be that the value of $\hat {p}$ is close to $p$?

## Case Study

A medical consultant helps guide transplant patients through all the stages of surgery. Out of a total of 62 clients, only 3 had complications:

$$
\hat{p} = 3/62 = 0.048
$$

Consultant claims that her rate is less than the national rate of 10%.

Question -- how confident can we be that $p$ is less than the national rate of 0.1?

## Distributions

We know that there's some **variation** in the data. But how much?

If this particular consultant had *another* 62 clients, how many of them might have complications?

\

This is similar to what we did last week.

We need to know the **sampling distribution**.

## Two Steps

1.  Find the distribution

2.  Use the distribution to find confidence interval

## Step 1 -- Find Sampling Distribution

-   Determine with math!\
    We'll come back to this in Chapter 13.

-   **Bootstrapping**!\
    Use the observed proportion to construct a new population and **resample** from that.

-   Someone else gives this to us. (e.g. see Chapter 12 homework)

## Bootstrap

Want 62 new outcomes each with a 0.048 chance of having a complication. Put 62 marbles in a bag

-   3 labeled "C"
-   59 labeld "NC"

1.  pick a marble
2.  note outcome
3.  put back into bag (sample with replacement)
4.  repeat

## Resample #1

"NC", "NC", "NC", "NC", "NC", "C", "NC", "C", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "C", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "NC", "C", "NC", "NC", "NC", "C", "NC", "NC"

\

Complications: 5/62 = 0.081

## Resample many times to get sample distribution

```{r}
#| echo: true

set.seed(25) 

boot_dist_donor <- organ_donor |>
  specify(response = outcome, success = "complication") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop")
```

## Visualize

```{r}
#| echo: true

boot_dist_donor |>
  ggplot( aes(x = stat)) +
  geom_histogram(binwidth = 0.011, col="white")
```

## Visualize

```{r}
#| echo: true

visualize(boot_dist_donor)
```

## Step 2 -- Use Sampling Distribution to get confidence interval

Estimate 95% confidence interval from graph

```{r}
visualize(boot_dist_donor)
```


## 95% confidence interval
```{r}
ci <- boot_dist_donor |>
  get_confidence_interval(level = 0.95)
```

```{r}
#| echo: false

visualize(boot_dist_donor) +
  shade_confidence_interval(ci)
```

## Confidence Interval #2

```{r}
#| echo: true

boot_dist_donor |>
  summarize(
    lower = quantile( stat, 0.025),
    upper = quantile( stat, 0.975)
  )
```

## Confidence Interval #3

```{r}
#| echo: true

ci <- boot_dist_donor |>
  get_confidence_interval(level = 0.95)

ci
```

## visualize

```{r}
#| echo: true

visualize(boot_dist_donor) +
  shade_confidence_interval(ci)
```

## Interpretation

We are 95% confident that the true proportion is between 0.0 and 0.113.

Notice that this interval contains the national rate of 0.1.

Therefore we cannot be confident (at a 95% level) that true proportion is less than 10%!
