---
title: "Lab 03 -- Voter Turnout"
categories: 
  - Lab
  - Suggested answers
fig-width: 6
fig-asp: 0.618
execute:
  echo: true
  
number-sections: true
number-depth: 1  
---

The goal of this lab is to demonstrate understanding of material from Chapters 11-13.  In particular, we will do an inference on a single population proportion and

- Critique data-based claims
- Compute and interpret confidence intervals.
- Articulate the basic steps in a hypothesis test.

# Introduction

In a [survey conducted by Survey USA](https://www.surveyusa.com/client/PollReport.aspx?g=4d09191d-b0b5-42cd-91c8-bd1bdf587199) between September 30, 2023 and October 3, 2023, 2759 registered voters from all 50 US states were asked:

> America will hold an election for President of the United States next November. Not everyone makes the time to vote in every election. Which best describes you? Are you certain to vote? Will you probably vote? Are the chances you will vote about 50/50? Or will you probably not vote?

Based on these responses, we can ask whether we expect voter turnout to be greater than it was in the 2020 presidental election.  That year, 66% of the voting-eligible population turned out for the presidential election (the highest rate for any national election since 1900).

## Packages and Data

For this analysis, we'll use the **tidyverse** and **tidymodels** packages.

```{r}
#| message: false

library(tidyverse)
library(tidymodels)
```

The data from this survey is in the file `voting-survey.csv` saved in your posit.cloud workspace.
```{r}
#| message: false
voting_survey <- read_csv("voting-survey.csv")
```

# Preliminary Analysis

::: {.callout-note title="Task 1" icon=false}
Make a bar chart to visualize the distribution of responses. Then calculate the proportion of respondents who are certain to vote in the next presidential election.
:::

```{r}
ggplot(voting_survey, aes(x = vote)) +
  geom_bar(fill = "cadetblue4")

voting_survey |>
  count(vote)
```

```{r}
voting_survey |>
  count(vote) |>
  mutate(prop = n/sum(n))

voting_survey |>
  count(vote) |>
  mutate(p_hat = n / sum(n)) |>
  filter(vote == "Certain to vote")
```



::: {.callout-tip title="Task 1 Solution" icon=false}
The proportion of survey respondents that say they are certain to vote is $\hat{p} = 0.696$

:::


## Estimation with Confidence Interval

Our goal in this section is to estimate the true proportion of registered US voters who are **certain to vote** in the next presidential election, based on these data.

::: {.callout-note title="Task 2" icon=false}
What is the parameter of interest?
:::

::: {.callout-tip title="Task 2 Solution" icon=false}

*Proportion of all registered voters, $p$, who are certain to vote in the next presidential election*
:::


We can use the bootstrap technique of Chapter 12 (and which we used in AE 07) to estimate a 95% confidence interval.  Below is a histogram of the sample distribution, along with the confidence interval shaded in. 




```{r}
#| label: bootstrap-ci
#| echo: false

voting_survey <- voting_survey |>
  mutate(vote = if_else(vote == "Certain to vote", "Certain to vote", "Not certain to vote"))

obs_stat <- voting_survey |>
  specify(response = vote, success = "Certain to vote") |>
  calculate(stat = "prop")

set.seed(1234)

boot_dist <- voting_survey |>
  specify(response = vote, success = "Certain to vote") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop")

ci <- boot_dist |>
  get_ci()

visualize(boot_dist) +
  shade_ci(ci)

# print(ci)
```




::: {.callout-note title="Task 3" icon=false}
Suppose the bounds of this confidence interval are L = 0.678 and U = 0.714.  Your friend offers the following interpretation:

> 95% of the time, the true proportion of registered US voters who are certain to vote in the next presidential election is between L and U.

Comment on this interpretation. Is it correct? If not, how would you fix it?
:::

::: {.callout-tip title="Task 3 Solution" icon=false}

*Not correct. We are 95% confident that 68.1% to 71.3% of all registered voters are certain to vote in the next presidential election.*

:::



# Hypothesis Testing

A newspaper claims that based on this study, the number of voters certain to vote in the 2024 election will be greater than it was in 2020.  Do these data provide convincing evidence for this claim?

::: {.callout-note title="Task 4" icon=false}
What are the null and alternative hypotheses?

:::

::: {.callout-tip title="Task 4 Response" icon=false}

*H0: There is nothing going on, 66% of registered US voters are certain to vote in the next presidential election. p = 0.66.*

*HA: There is something going on, more than 66% of registered US voters are certain to vote in the next presidential election. p \> 0.66.*

:::


::: {.callout-note title="Task 5" icon=false}
Before we go any further, we should decide on a discernment level.  What discernment level will you use and how did you decide? In your response, you should say something about Type I and Type II errors.

:::

::: {.callout-tip title="Task 5 Response" icon=false}
*Type I error -- incorrectly reject the null hypothesis.  This means we incorrectly conclude that more than 66% will vote.  Consequences might include a false sense of vote engagement with candidates or issues.  This could lead to complacency around get out the vote efforts.  If we want to avoid Type I errors, we might set $\alpha = 0.01$*

*Type II error -- incorrectly failing to reject the null hypothesis.  This means incorrectly concluding that we don't have evidence to suggest voter turnout will be higher.  Consequences might include underestimating voter engagement with candidates or issues.  If we want to avoid Type II errors, we might set $\alpha = 0.1$*

:::



::: {.callout-note title="Task 6" icon=false}
Because the conditions of the CLT are met, we can use a normal distribution to find the p-value instead of using a simulation.  Assuming the null hypothesis is true, where would this normal distribution be centered?  Replace the 0 in the code chunk below with the correct value, and replace the 1 with 0.01 which is the appropriate standard error (SE) for this problem.  We will see a formula that gives us this value in Chapter 17 but we'll just assume this for now. 

:::

::: {.callout-tip title="Task 6 Response" icon=false}
*The null hypothesis is that the proportion has not changed - $p = 0.66$ - so distribution would be centered there. *

:::


```{r}
openintro:: normTail( m = 0.66, s = 0.01)
```

 
::: {.callout-note title="Task 7" icon=false}
Using the SE value 0.01 and appropriate the values for $\hat{p}$ and $p$, find the Z-score for the observed statistic. 

Then find the p-value corresponding to this Z-value.
:::

::: {.callout-tip title="Task 7 Response" icon=false}
Enter your responses to Task 7 here.
*The Z-score is approximately 3.6.*

```{r}
1-pnorm(3.6)
```

*The p-value is approximately 0.0002, which is much lower than even the 1% discernability level, so we would reject the null hypothesis. The data provide convincing evidence for the alternative hypothesis.*

:::


### Interpreting the p-value

::: {.callout-note title="Task 8" icon=false}
Suppose the p-value you found is P, and your friends are in disagreement about the interpretation about this value. One friend claims:

> The probability that 66% of all of registered US voters are certain to vote in the next presidential election is approx P.

Another friend claims:

> The probability that more than 66% of all of registered US voters are certain to vote in the next presidential election is approx P.

Who is right? Explain your reasoning.
:::

::: {.callout-tip title="Task 8 Response" icon=false}
Enter your responses to Task 8 here.

*Both friends are wrong. The p-value is neither the probability of $H_0$ being true, nor the probability of $H_A$ being true.*

:::


::: {.callout-note title="Task 9" icon=false}
Summarize your findings in the context of this problem.
:::

::: {.callout-tip title="Task 9 Response" icon=false}
Enter your response to Task 9 here.


:::

