---
title: Inference for Single Proportion - part 2
subtitle: Chapter 16
author: ""
format: revealjs
auto-stretch: false
editor: 
  markdown: 
    wrap: 72
---


## Medical consultant example...

```{css, echo = FALSE}
.red {
color: #B8001F;
}
```

A consultant tried to attract patients by noting the average complication rate for liver donor surgeries in the US is about 10%, but her clients have had only 3 complications in the 62 liver donor surgeries she has facilitated. She claims this is strong evidence that her work meaningfully contributes to reducing complications (and therefore she should be hired!).

$$
\hat{p} = \frac{3}{62} = 0.048
$$

## State the null hypothesis

Claim is that $\hat{p}$ is [less]{.red} than then national average of 0.1

- $H_0: p = 0.1$
- $H_A: p < 0.1$

## Discernment level?

- Type I error:  mistakenly reject $H_0$ when it's true

  + Misled by false advertising!
  + Make $\alpha$ smaller to avoid

- Type II error:  mistakenly fail to reject $H_0$ when it's false

  + Being cautious leads to missing out on above average care.
  + Make $\alpha$ larger to avoid




## Check conditions for CLT

- Independent? yes

- Large enough sample size? [no!]{.red}

    $n p_0 = 62 \cdot 0.1 = 6.2$
    
    $n (1-p_0) = 62 \cdot 0.9 = 55.8$
    
\

[What happens if we ignore this?]{.red}

## Calculate Z

$$
SE = \sqrt{ \frac{p_0 (1-p_0)}{n}} =  \sqrt{\frac{0.1 \cdot 0.9}{62}} = 0.038
$$
$$
Z = \frac{\hat{p} - p_0}{SE} = \frac{ 0.048 - 0.1}{0.038} = -1.37
$$

## Find p-value for Z = -1.37

```{r}
#| echo: false
openintro::normTail(m=0, s=1, L=-1.37)
```
```{r}
pnorm(-1.37)
```

## Compare to discernment level

Suppose $\alpha = 0.1$ (trying to avoid Type II errors)

\

Since p-value (0.085) is [less]{.red} than $\alpha$, we [do]{.red} have evidence to reject the null hypothesis.

\

We think that our consultant's complication rate is significantly below the national average.


:::{.fragment .center-align}
[BUT WAIT!]{.red}

[CLT isn't valid!!]{.red}
:::

---
### Do a simulation!!

```{r}
#| echo: false
#| message: false

library(infer)
library(tidyverse)

organ_donor <- read_csv("organ-donor.csv")
```

```{r}
set.seed(2024)

null_dist <- organ_donor |>
  specify(response = outcome, success = "complication") |>
  hypothesize(null = "point", p = 0.1) |>
  generate(reps = 1000, type = "draw") |>
  calculate(stat = "prop")
```

## Simulated Null-Distribution

```{r}
visualize(null_dist) +
  shade_p_value(obs_stat = 0.0484, direction = "less")
```

## Actual p-value?

```{r}
get_p_value(null_dist, obs_stat = 0.0484, direction = "less")
```

\

- [Not]{.red} less than $\alpha = 0.1$
- Previously had found $p = 0.085$ 
- conclusion depends on discernment level as well as how we found p-value

