---
title: Exploring numerical data
subtitle: Chapter 5
format: revealjs
auto-stretch: false
---

## In groups

-   Today's homework: section 5.10, #1, 2, 5a,b


## Scatterplots

Compare **two** numerical variables

```{r}
#| echo: false
#| message: false
library(palmerpenguins)
library(tidyverse)
```

```{r}
#| echo: false
#| warning: false

ggplot(data = penguins, 
       mapping = aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()

```

## Dot Plot

Visualize the *distribution* of **one** numerical variable
```{r}
#| warning: false
ggplot(data = penguins, 
       mapping = aes(x = flipper_length_mm)) +
  geom_dotplot(dotsize=0.75, binwidth=1)
```

## flipper_length_mm

```{r}
penguins |> count(flipper_length_mm) |> print(n=20)
```


## Mean (average) flipper length?
```{r}
mean(penguins$flipper_length_mm, na.rm = TRUE)
```

\

```{r}
penguins |>
  summarize( mean = mean(flipper_length_mm, na.rm = TRUE))
```

## Visualize the Mean

A measure of **center** of a distribution.

```{r}
#| warning: false
#| echo: false

ggplot(data = penguins, 
       mapping = aes(x = flipper_length_mm)) +
  geom_dotplot(dotsize=0.75, binwidth=1) +
  geom_point(aes(x=201, y=-0.05), shape = 24, size = 5, fill="red")

```



## Calculuating the Mean

Sample mean $\overline{x}$

$$
\overline{x} = \frac{ x_1 + x_2 + x_3 + \ldots + x_n}{n}
$$

\

Population mean $\mu$ (Greek letter "mu")
$$
\overline{x} \approx \mu
$$

\
**Question:** what is the average age of the people in this room?

## Histogram

```{r}
#| warning: false
#| fig-width: 8
#| fig-asp: 0.5
#| fig-align: center

ggplot(data = penguins, 
       mapping = aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 5, color="white" )
```

This distribution is **bimodal** and **right skewed**

## Body Mass
```{r}
#| warning: false
#| fig-width: 8
#| fig-asp: 0.5
#| fig-align: center

ggplot(data = penguins, 
       mapping = aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, col="white")
```

This distribution is **unimodal** and **right skewed**

## Bill Depth
```{r}
#| warning: false
#| fig-width: 8
#| fig-asp: 0.5
#| fig-align: center

ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm)) +
  geom_histogram(binwidth = 0.5, col="white")
```

This distribution is **bimodal** and **right skewed**

## Mean (average) Body Temperature

What do you think it is?


## Mean (average) Body Temperature
```{r}
#| echo: false
load("thermometry.rda")
```

```{r}
mean(thermometry$body.temp)
```

```{r}
#| message: false
#| warning: false
#| echo: false
ggplot(data = thermometry, 
       mapping = aes(x = body.temp)) +
  geom_histogram(binwidth = 0.4, col="white") +
  geom_point(aes(x=mean(body.temp), y=-0.05), shape = 24, size = 5, fill="red")
```

## Variance

Measure of *variation* or how *spread out* distribution is.  It's the *average squared distance from the mean*.

Sample **variance** is $s^2$

where *s* is the sample **standard deviation**
$$
s = \sqrt{ \frac{ \sum_{i=1}^n (x_i - \overline{x})^2 }{n-1}}
$$

\
\

Population variance:  $\sigma^2$
(Greek letter "sigma")


## Standard Deviation

*s* is the sample **standard deviation**.  Represents the typical deviation from the mean
$$
s = \sqrt{ \frac{ \sum_{i=1}^n (x_i - \overline{x})^2 }{n-1}}
$$

### Empirical Rule
Typically, about 68% of the data (observations) lie within one 
s.d. of the mean.

About 98% of the data lie within two s.d. of the mean.

These percentages are **not** hard and fast rules!

## Body Temperature

```{r}
thermometry |>
   summarize( mean = mean(body.temp), sd = sd(body.temp))
```

Using the **empirical rule**, about 68% of observations lie in what range of temperatures?

