---
title: Inference for Single Proportion
subtitle: Chapter 16
author: ""
format: revealjs
auto-stretch: false
editor: 
  markdown: 
    wrap: 72
---

## Looking Ahead

```{r}
#| echo: false
#| message: false

library(openintro)
```

### We will take a closer look at [inference methods]{.red} in different contexts:

-   Ch 16: Single Proportion
-   Ch 17: Two Proportions
-   Ch 18: Two-way tables (multiple proportions)
-   Ch 19: Single Mean

## Recall our Hypothesis Testing Framework

```{css, echo = FALSE}
.red {
color: #B8001F;
}
```

1.  State the [null]{.red} and the [alternative]{.red} hypotheses

2.  Choose a [sample]{.red}, collect and analyze the data

3.  How [likely]{.red} is it to see data like what we observed, [if the
    null hypothesis were true]{.red}?

4.  If very unlikely, we [reject]{.red} the null hypothesis. Otherwise,
    we [cannot reject]{.red}.

## p-value

-   p-value is a [probability]{.red} -- how likely is it to see our
    observation **if the null hypothesis is true**?

-   Finding the p-value

    -   If conditions for Central Limit Theorem [are met]{.red}, can use
        a [mathematical model]{.red}, i.e. normal curve - find z-scores,
        etc.
    -   If conditions for Central Limit Theorem [are not met]{.red}, use
        [simulation methods]{.red}, i.e. bootstrapping or randomization

## Mathematical Model

### Needed Conditions

The sampling distribution $\hat{p}$ based on a sample size $n$ from a
population with a true proportion $p$ is nearly normal if:

1.  Sample's observations are [independent]{.red} - roughly:
    observations do not impact another observations

2.  [Success-failure condition]{.red} -- we expect to see at least 10
    successes and 10 failures.

    -   $np \geq 10$ and $n(1-p) \geq 10$

## Standard Error Formula (new)

If conditions are met then distribution of $\hat{p}$ is nearly normal
with variability given by the standard error

$$
SE = \sqrt{ \frac{p(1- p)}{n}}
$$ Note we (usually) don't know $p$ (it's the population parameter) so
we use our best guess for $p$ intead:

-   For hypothesis testing, use $p_0$ (null hypothesis)
-   For confidence intervals, use $\hat{p}$

## Confidence Interval

$$
\mbox{point estimate} \pm z^\ast * SE
$$

$z^\ast * SE$ is the [margin of error]{.red}

$z^\ast$ is determined by the desired confidence level.



## Confidence Level

![](images/confidence_levels){width="100%"}

95% confidence level means $z^\ast = 1.96$



## Example: confidence interval

::: small
> In survey of 826 Portland residents, 48.3% of respondents say that
> homelessness is Portland's greatest challenge.
:::

-   Are CLT conditions met?
    -   Are observations independent? Yes - random sample
    -   Success/failure? Yes
        -   $np = 3290*0.483 = 1589 \geq 10$
        -   $n(1-p) = 3290 * (1 - 0.483) = 1700 \geq 10$

## Example: confidence interval

::: small
> In survey of 3290 Portland residents, 48.3% of respondents say that
> homelessness is Portland's greatest challenge.
:::

\

Standard error? 
$$
    SE = \sqrt{ \frac{p(1- p)}{n}} = \sqrt{ \frac{ 0.483(1-0.483)}{3290}} = 0.009
$$

## Example: confidence interval

::: small
> In survey of 3290 Portland residents, 48.3% of respondents say that
> homelessness is Portland's greatest challenge.
:::

We are 95% confident that the true proportion of Portlanders who agree that homelessness is the city's greatest challenge

$$
48.3 \pm 0.9\%
$$

## Hypothesis Test for single proportion

Z-score tells us how the sample proportion ($\hat{p}$) differs from the hypothesized proportion ($p_0$)

$$
Z = \frac{ \hat{p} - p_0}{SE} = \frac{ \hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}
$$

## Example

> Supporters of ballot measure that would impose stricter regulations on payday lenders commissioned a survey that asked 826 random voters if they supported such a measure.  The survey found that 51% of respondents said they did support it.  Is there convincing evidence to say that voters support this measure?

## Example:  hypotheses

Null hypothesis:  there is no support 
$$
p_0 = 0.5
$$

Alternative hypothesis:  there is support
$$
p_0 \geq 0.5
$$

## Example: conditions

+ independent
+ success/failure
  - $np = 825*0.5 = 413 \geq 10$
  - $n(1-p) = 825*0.5 = 413 \geq 10$
  
## Example: standard error

$$
SE = \sqrt{ \frac{ p_0(1-p_0)}{n}} = \sqrt{ \frac{0.5(1-0.5)}{826}} = 0.017
$$


```{r out.width="80%"}
#| echo: false
#| warning: false


openintro::normTail(m = 0.5, s = 0.017, U = 0.51 )
```

## Example: Z-score

$$
Z = \frac{ \hat{p} - p_0}{SE}  = \frac{ 0.51 - 0.50}{0.017} = 0.59
$$


```{r out.width="80%"}
#| echo: false
#| warning: false

openintro::normTail(m = 0, s = 1, U = 0.59 )
```

## Example:  find p-value

```{r out.width="60%"}
#| echo: false
#| warning: false

openintro::normTail(m = 0, s = 1, U = 0.59 )
```


```{r}
1 - pnorm(0.59, mean = 0, sd = 1 )
```
Because p-value is larger than 0.05, we do not reject $H_0$.  There is [not]{.red} convincing evidence that there is support for this measure.


## What if conditions for CLT are not met?

+ Independence?  

  need more advanced methods (beyond scope of our class)

+ Small sample size?  

  Use [simulations]{.red} -- next time!


```{r}
```
